---
title: "Analysis of VLW 2Dvs3D combined data, part 2"
output: 
    html_notebook:
        self_contained: yes
author: "Darren R Tyson"
date: "2021-02-15"
---

### Analysis of dead cell fractions
In part 1 (`VLW_V23_combined_analysis.Rmd`), all data were organized to include drug annotation and image file paths, in addition to cell count and ch2-pos counts. Use these data to examine time and drug concentration effects on cell death kinetics.

```{r Setup}
library(diprate)

OVERWRITE <- FALSE
```

For these analyses we do not need `file_name`
```{r Load data}
d <- read.csv("../data/VLW_V23ab_dataset.csv", as.is=TRUE)
d <- d[order(d$uid,d$time),]
d$file_name <- NULL
```

```{r Prep counts by cell line, include=FALSE}
pd <- prepCountByCL(d[grepl("V2",d$plate.name),])
```

Examine the data from the previously identified hits. Focus on two cell lines. 16T best-looking 3D structures, 21T most clinically relevant, but worst looking and slowest growing.

* pralatrexate (low potency) and ixazomib (high potency and efficacy) on 16T and 21T.

```{r 2D data subset, message=FALSE, warning=FALSE}
mydrugs <- c("bortezomib", "ixazomib", "YM155", "Zibotentan", "Romidepsin", "cabazitaxel", "delanzomib",
             "ispinesib", "KX2391", "triptolide", "carfilzomib", "rigosertib", "pralatrexate")
mycl <- c("16T","21T")
s2d <- lapply(mycl, function(cl) d[d$drug1 %in% mydrugs & 
                                      d$cell.line==cl & 
                                      grepl("V2",d$uid),])
```

Save drug subsets into objects.

```{r 2D data, message=FALSE, warning=FALSE}
a <- subset(s2d[[1]], drug1 %in% mydrugs)
b <- subset(s2d[[2]], drug1 %in% mydrugs)
```


```{r 3D data subset, message=FALSE, warning=FALSE}
s3d <- lapply(mycl, function(n) d[d$drug1 %in% mydrugs & 
                                      d$cell.line==n & 
                                      grepl("V3",d$uid),])
```

#### Screen 3D data by cell count > 40
Many 3D cultures had few, if any, cells in the field of view. Must exclude.
```{r 3D data}
a3 <- subset(s3d[[1]], drug1 %in% mydrugs & orig.cell.count >= 40)
b3 <- subset(s3d[[2]], drug1 %in% mydrugs & orig.cell.count >= 40)
```

### 16T 2D
Examine the death kinetics (fraction of ch2 positive cells over time).
```{r 16T 2D, fig.height=5, fig.width=8}
plotAllCh2(list("16T"=a), count_cn="orig.cell.count")
```

### 16T 3D
```{r 16T 3D, fig.height=5, fig.width=8}
plotAllCh2(list("16T"=a3), count_cn="orig.cell.count")
```
### 21T 2D
```{r 21T 2D, fig.height=5, fig.width=8}
plotAllCh2(list("21T"=b), count_cn="orig.cell.count")
```

### 21T 3D
```{r 21T 3D, fig.height=5, fig.width=8}
plotAllCh2(list("21T"=b3), count_cn="orig.cell.count")
```


### Apply simple logistic model to death kinetics data

```{r logistic function}
myl3 <- function(x, min=0,max=0.5,mid_x=48, dr=0.05) max/(1 + exp(-dr*(x-mid_x))) 
```


```{r test plot of logistic function}
curve(myl3, from=0, to=120)
```

### Use DRC 3-param logistic function to fit data
The `drc` library has a number of dose--response model fitting algorithms. The `L.3` model is a 3-parameter logistic (not log-logistic) model that can be applied to the time course data.  

Test this with a subset of the data (21T + delanzomib).

The model is trying to explain the change in dead cell fraction over time.

```{r delanzomib}
mydat <- b[b$drug1=="delanzomib",]
test.m <- drc::drm(ch2.pos/orig.cell.count ~ time, factor(signif(log10(drug1.conc),3)), 
                   data = mydat, fct = drc::L.3())

plot(test.m, legendPos=c(20,0.85), log="")
```
### Extract maximum dead cell fraction from curves
Pull from fit coefficients (upper bound; coefficient "d").
```{r Max dead cell fraction}
myconc <- names(coef(test.m))[grep("d",names(coef(test.m)))]
myconc <- gsub("d:","",myconc)
myconc <- as.numeric(myconc)

plot(myconc, coef(test.m)[grep("d",names(coef(test.m)))], ylab="Max dead fraction", main="delanzomib")
```
### Examine an different drug
Determine whether predicted time course trajectories match those of the ixazomib-treated cells.
```{r ixazomib}
mydat <- b[b$drug1=="ixazomib",]
test.m <- drc::drm(ch2.pos/orig.cell.count ~ time, factor(signif(log10(drug1.conc),3)), 
                   data = mydat, fct = drc::L.3())

plot(test.m, legendPos=c(20,0.85), log="")
```

```{r Plot ixazomib DRC}
plot(myconc, coef(test.m)[grep("d",names(coef(test.m)))], ylab="Max dead fraction", main="ixazomib")

```

This looks like it is working appropriately and the resultant dose--response curve could be fit by a LL4 model (lower bound should not be fixed to 0). However, the model fit values currently extend beyond the measured time range (e.g., Einf) and may not be accurate. A safer, albeit less elegant, way would be to simply capture the largest death fraction across the entire time course for each unique condition and plot those as the effect metric.

```{r max death fraction}
death_frac_max <- sapply(unique(d$uid), function(id) 
{
    z <- d[d$uid==id,]
    z <- z[z$orig.cell.count !=0,]
    out <- max(z$ch2.pos/z$orig.cell.count)
    out[is.infinite(out)] <- NA
    return(out)
})
names(death_frac_max) <- unique(d$uid)

s <- d[!duplicated(d$uid),]

# all(s$uid == names(death_frac_max))

s$death.frac.max <- death_frac_max

# Add 2D or 3D to cell line name to more easily distinguish
s$cell.line <- paste0(s$cell.line,"_",substr(s$plate.name,2,2),"D")

# Make cell.count into surviving percentage
s$cell.count <- signif((1 - s$death.frac.max) * 100,3)

# make time all 72h (not correct or relevant!)
s$time <- 72
```

### Percent surviving dose--response curves
Using 1 - max death fraction as the effect metric, fit a 4-parameter log-logistic model (LL.4). 
```{r DFM DRC, warning=FALSE}
k <- s[s$plate.name=="V2-11T-D3" & s$drug1=="ixazomib",]
m <- drc::drm(cell.count ~ drug1.conc, data = k, fct = drc::LL.4())
plot(m, type="all", ylim=c(0,100), ylab="Percent surviving")
```

### Iterate over unique drugs and plates (cell line and condition)
```{r All drugs surviving fraction, message=FALSE, warning=FALSE}
alldrugs <- unique(d$drug1)[-1] # remove "control"
all.m <- lapply(alldrugs, function(drug)
{
    z <- s[s$drug1==drug,]
    ucl <- unique(z$cell.line)
    cld <- lapply(ucl, function(cl) 
    {
        dtf <- z[z$cell.line==cl,]
        out <- tryCatch({drc::drm(cell.count ~ drug1.conc, data = dtf, fct = drc::LL.4())},
                 error=function(cond) {NA})
        return(out) 
    })
    names(cld) <- ucl
    return(cld)
})

names(all.m) <- alldrugs
```

```{r plot DRC}
plotMultiCurve(unlist(all.m, recursive=FALSE)[9:16], ylim=c(0,100))
```

### Save surviving percentage data for Thunor
Still need unique plate id (`upid`) 
```{r Save surv for Thunor}
s$upid <- paste(s$expt.id,s$plate.name,sep="_")

# remove NAs
s <- s[!is.na(s$cell.count),]

fn <- "../data/VLW_001+V23b_SurvPct_Thunor_dataset.csv"

if(!file.exists(fn) | OVERWRITE) write.csv(s, file=fn, row.names=FALSE)
```

### Import Thunor curve-fit data
Data uploaded to Thunor (`VLW_SurvPercent`) were processed and the `Viability Parameters` were downloaded as .

```{r}
v <- read.csv('../data/VLW_SurvPercent_viability_params.tsv', sep="\t")
```

Order the drugs by `aa_obs` (highest to lowest) and save subset with `aa_obs < 0.5` to object `b`. Count how many of each cell line are found in this subset (looking for bias in cell lines or 2D/3D condition)
```{r}
v <- v[order(v$aa_obs, decreasing=TRUE),]
b <- v[v$aa_obs > 0.5,]
diprate::nEach(b$cell_line)
```
16T_3D looks overrepresented and 29T_3D looks underrepresented.

```{r}
b[b$cell_line=="29T_3D",]
```

### Sort values by drugs
Looks at sum of aa_obs as metric of overall effect.
Also get sd to assess similarity across cell lines and 2D/3D.
```{r}
k <- lapply(unique(v$drug), function(x) 
              v[v$drug==x,c("cell_line","aa_obs")])
names(k) <- as.character(unique(v$drug))

a <- sapply(names(k), function(x) sum(k[[x]]$aa_obs))
a <- sort(a, decreasing=TRUE)

# reorder k to match a
k <- k[names(a)]

std.dev <- sapply(names(k), function(x) sd(k[[x]]$aa_obs))
cv <- sapply(names(k), function(x) sd(k[[x]]$aa_obs)/mean(k[[x]]$aa_obs))

av <- list(orig.data=k, sum.aa_obs=a, std.dev=std.dev, cv=cv)

```

### Include DIP rate-based (observed) Emax values from 2D condition for each cell line
First, load all DIP rate-based parameters (DIP parameters downloaded from [DipDB](https://dipdb.lolab.xyz/dataset/193)). Extract the `emax_obs` values from each cell line for each drug as a new data.frame (saved in `emo` for emax_obs).

```{r}
d <- read.csv("../data/VLW_23ab_2D_only_dip_params.tsv", sep="\t", as.is=TRUE)

# emo = emax_obs
emo <- data.frame(do.call(rbind,lapply(unique(d$drug), function(x) d[d$drug==x,"emax_obs"])))
colnames(emo) <- paste0("emax_obs_",unique(d$cell_line))
rownames(emo) <- unique(d$drug)
```

Organize a data.frame to save for Kensey and Vivian.
```{r}
o <- data.frame(viability.aa_obs.sum=a, viability.aa_obs.std.dev=std.dev, viability.aa_obs.cv=cv)
if(!file.exists(fn) | OVERWRITE) write.csv(o, file="../data/max_death_DRC_params.csv", row.names=FALSE)
head(o)
```

