---
title: "Identify images for Kensey"
output: html_notebook
---
### For Kensey Bergdorf manuscript
Reviewers asked to see reference images in 3D of each cell line of the best-performing drugs.  

Specific information from Kensey: One of the reviewers has asked for images from the top 3 compounds (bortezomib, cabazitaxel, and YM155) across the four cell lines. I just need a composite image from high concentration and low concentration for each drug in each line, so 8 images total.


Choosing time points after 100 hours.
```{r}
d <- read.csv("../data/VLW_V23ab_dataset.csv", as.is=TRUE)
d <- d[order(d$uid,d$time),]

mydrugs <- c("bortezomib","cabazitaxel","YM155")
z <- d[d$drug1 %in% mydrugs & grepl("V3",d$plate.name),]

max_conc <- sapply(mydrugs, function(drug) max(d[d$drug1==drug,"drug1.conc"]))

z <- z[z$drug1.conc %in% max_conc,]
z <- z[z$time>100,]

z$ucond <- paste(z$cell.line, z$drug1, sep="_")
```

#### Determine which image for each unique condition has highest cell count
```{r}
unique(z$ucond)
```
Missing 29T_bortezomib (see below); get images with highest cell count for others.
```{r}
a <- do.call(rbind, lapply(unique(z$ucond), function(cond)
{
    temp <- z[z$ucond==cond,]
    temp[temp$orig.cell.count==max(temp$orig.cell.count),]
}))
```

### Need other time points or conc
Bortezomib in 29T was missing

Needed to find an earlier time point (prior to 100 h)
```{r}
b <- d[d$drug1=="bortezomib" & d$drug1.conc %in% max_conc & d$cell.line=="29T" & grepl("V3", d$plate.name) & d$time>72,]
b$ucond <- paste(b$cell.line,b$drug1,sep="_")
```

```{r}
a <- rbind(a,b[1,])
```

Get control wells that correspond to the same times.

```{r}
ctrl <- do.call(rbind, apply(a[,c("plate.id","time")], 1, function(x) d[d$plate.id==x["plate.id"] & d$time==x["time"] & d$drug1.conc==0,]))
ctrl$ucond <- paste(ctrl$cell.line, ctrl$drug1, sep="_")
ucond_ctrl <- unique(ctrl$ucond)
temp <- do.call(rbind, lapply(ucond_ctrl, function(cond) 
{
    x <- ctrl[ctrl$ucond==cond,]
    x[x$orig.cell.count==max(x$orig.cell.count),]
}))

ctrl <- temp[!duplicated(temp$plate.id),]
```

Append control wells to drug treated wells
```{r}
a <- rbind(a,ctrl)
a <- a[order(a$ucond),]
rownames(a) <- NULL
```

### Show data to be saved
```{r}
a
```

### Write file
```{r}
write.csv(a, file="../data/VLW_images_Bergdorf_pub.csv", row.names=FALSE)
```

